Complexity-stability relationships in competitive disordered dynamical systems

### 2024-09-17 papers/response-referee/old/response-referee.md

# Response
## Referee A

We appreciate the positive feedback from the referee. We address the remaining comment below.

*In my opinion, the arguments leading to Eq. (11) and the discussion*  
*after this equation are not very clear. What is the solution to the*  
*consistency condition for <x_^gamma> and <x_^2gamma>? Do these two*  
*quantities depend on N, and if so, how?*

*For example, for the case gamma=1, discussed after Eq. (11), both*  
*means correspond exactly to the moments <x_> and <x_^2> and, as the*  
*authors noted, are not well defined for the probability distribution*  
*(11). This means that conclusions need to be drawn more carefully.*

We have now included the analysis on the cut-off argument previously in the SI in as an appendix in the main text, as we changed the format from letter to regular article. 
Moreover, we have now modified the paragraph below Eq. (11) and conclude with a sentence clarifying that the same "renormalization" procedure we describe for $\alpha = 1$, $\beta = 2/3$, and $\gamma = 1$ can be extended to other choices of exponents in the case they result in diverging expectation values.
Finally, for clarity, we now write explicitly the self-consistent equations for $\Lambda$, $\langle x_{*}\rangle_{\Lambda}$, and, $\langle x_{*}^2\rangle$.

*The authors clarify most of my previous concerns and clarify the*  
*scaling issue, and the paper is surely improved. I have still one*  
*relevant concern, I hope the authors fix before publication.*

We thank the referee for the previous comments, which helped improving the paper, and address the last concern below.

- *In the response to reviewers the authors claim they better highlight*  
	*that their result are valid only for competitive Lotka Volterra. It is*  
	*true they added a couple of times the word competitive but in my*  
	*opinion the paper is not clear enough on this points and the results* *are presented are too general.*

1) *The title "Beyond May: Complexity-stability relationships in*  
	*disordered dynamical systems" seem to present results that are* 
	*completely general. I suggest to make a more proper title such as*  
	*""Beyond May: Complexity-stability relationships in competitive*  
	*disordered Generalized Lotka Volterra Dynamics", or something similar.*  
	*Otherwise the title is misleading.*

	 We have now modified the title to make it clear that our work regards competitive disordered interactions only. We have additionally modified the abstract to stress our focus on competitive systems "We focus on competitive systems, and, using ...".

2) *Many comparison are made with respect to systems that have not such*  
	*constraint (positive mu), and this makes confusion.*

	*E.g. 1, " the classic generalized Lotka-Volterra (GLV) competition*  
	*model studied by Bunin and collaborators [25, 26] corresponds to f, g,*  
	*h all linear." But this is not true as in Bunin and Biroli works mu*  
	*can also be negative, is not constrained to assume positive value.*

	We now modified this sentence specifying that the linear case of our model corresponds to GLV only when constrained to competitive interactions.

	*E.g. 2, "Compared to the case where μ and σ scale with N (considered*  
	*in [25] and follow-up works), our choice reflects a strong-interaction*  
	*regime; it has been employed, for example, to study ecological*  
	*scenarios with large differences in species abundance and chaotic*  
	*turnover [35]". Again in the cited papers mu can be also negative.*

	We have removed the reference to [25] in order to avoid confusion. 
	We kept reference [35], because in this paper only competitive interactions ($\mu > 0$) are considered, as in our work.

- *I'm not sure it is worth to have a supplementary file. If it is not*  
	*published as letter and thus there is room it would be suitable to*  
	*better explain already in the main text the finite size argument to*  
	*solve self consist equations and the sensitivity to random exponents*

	*Justification for/against Letter:*

	*These calculations hold for a very specific cases and makes many*  
	*assumption that in general do not hold. Having more space the authors*  
	*could include the SI in the main text.*

	Following the indication of the referee, we have now included the SI in the main text and changed the format from Letter to Regular article.

---
# OLD
## Referee A

### 2024-09-17 papers/submission/beyond-may.tex
minor edits
minor corrections

\documentclass[
 prl,
 twocolumn,
 amsmath,
 amssymb,
 aps,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\usepackage{notes2bib}
\usepackage{hyperref}
\newcommand{\red}{\textcolor{red}}
\usepackage[justification=raggedright]{caption}

@@

\title{Complexity-stability relationships in disordered dynamical systems}

@@

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
\end{abstract}

@@

%Introduction
\paragraph*{}
\emph{Introduction.---}Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is, on the face of it, a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology be studying in the first place? 
In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022,Hatton2024}.

@@

%Model
\paragraph*{}
\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}>0$, such that it implies a negative effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
We refer to $f$ as the ``production function'' and $g$ as the ``coupling function''.

This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
In particular, the classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, at the individual organism level (populations of cells), growth has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.

%Homogeneous interactions
\paragraph*{}
\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu_s - \mu) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
an elementary calculation shows that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions. This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and coupling functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).

@@

%Random interactions: stability condition
\paragraph*{}
\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.

@@

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution that describes the ensemble of the stationary solutions of the system by means of a single representative random variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \mu N \langle x_*^{\gamma}\rangle + \sigma\sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}. Compared to the case where $\mu$ and $\sigma$ scale with $N$ (considered in \cite{bunin2017ecological} and follow-up works), our choice reflects a strong-interaction regime; it has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}. As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.

The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2N \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \mu N \langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently. Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
This problem can be overcome by considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$, where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$, see Supplemental Material.

Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently: $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$. The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}. Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

@@

Finally, we check the robustness of our findings in the scenario in which the exponents are not the same for every variable $x_i$. As a test, we can sample exponents from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$. 
Simulations show that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$ (Supplemental Material).

%Conclusion
\paragraph*{}
\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
In this Letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where self-interactions grow slower than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

@@

>>>>

\documentclass[
 pre,
 twocolumn,
 amsmath,
 amssymb,
 aps,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\usepackage{notes2bib}
\usepackage{hyperref}
\newcommand{\red}{\textcolor{red}}
\usepackage[justification=raggedright]{caption}
\renewcommand{\appendixname}{APPENDIX}

@@

\title{Complexity-stability relationships in competitive disordered dynamical systems}

@@

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in light of recent developments in both ecology and random matrix theory. 
    We focus on competitive systems, and, using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
\end{abstract}

@@

%Introduction
\section{Introduction}
Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is, on the face of it, a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology be studying in the first place? 
In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022,Hatton2024}.

@@

%Model
\section{Model}
Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}>0$, such that it implies a negative effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
We refer to $f$ as the ``production function'' and $g$ as the ``coupling function''.

This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
In particular, the classic generalized Lotka-Volterra (GLV) model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally}, when constrained to competitive interactions, corresponds to $f$, $g$, $h$ all linear. However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, at the individual organism level (populations of cells), growth has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.

%Results
\section{Results}
%Homogeneous interactions
\subsection{Homogeneous interactions}
Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu_s - \mu) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
an elementary calculation shows that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions. This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and coupling functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).

@@

%Random interactions: stability condition
\subsection{Random interactions: stability condition}
We now consider the case of random interactions.
Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.

@@

%Random interactions: equilibrium distribution
\subsection{Random interactions: equilibrium distribution}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution that describes the ensemble of the stationary solutions of the system by means of a single representative random variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \mu N \langle x_*^{\gamma}\rangle + \sigma\sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}. Compared to the case where $\mu$ and $\sigma$ scale with $N$, our choice reflects a strong-interaction regime; it has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}. As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.

The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2N \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \mu N \langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently.

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the system is always feasible, and the distribution can be normalized. 
However, the first and second moments formally diverge.
This problem can be overcome by considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$, where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$, i.e., such that there is statistically less than 1 variable out of $N$ with value above $\Lambda$; see Appendix~\ref{appendix a}.

Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}$:
\begin{align}
    \int_{0}^{\Lambda}dx_*P(x_*) &= 1 - \frac{1}{N} , \\    
    \langle x_* \rangle_{\Lambda} &= \int_0^{\Lambda}dx_*x_*P(x_*) , \\
    \langle x_*^2\rangle_{\Lambda} &= \int_0^{\Lambda}dx_*x_*^2P(x_*) .
\end{align}
The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}. 
Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error. Notice that the same ``renormalization'' procedure described above can be extended to other choices of the exponents $\alpha$, $\beta$, and $\gamma$, if needed, to compute $\langle x_*^{\gamma}\rangle$, and $\langle x_*^{2\gamma}\rangle$.

Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.

@@

Finally, we check the robustness of our findings in the scenario in which the exponents are not the same for every variable $x_i$. As a test, we can sample exponents from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$. 
Simulations show that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$ (Appendix~\ref{appendix b}).

%Discussion
\section{Discussion}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
In this Letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where self-interactions grow slower than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

@@

\appendix

\section{CUT-OFF}
\label{appendix a}
In this appendix we consider a case amenable to analytical treatment to exemplify the argument behind the cut-off $\Lambda$ introduced in the main text to deal with diverging moments distributions.

Consider the case of a power law distribution 
\begin{equation}
  P(x)=\frac{x^{-\beta}}{\mathcal{Z}} ,
\end{equation}
defined from 1 to $\infty$ and with 
\begin{equation}
  \mathcal{Z}=\int_1^{\infty}dxx^{-\beta}.  
\end{equation}

Let us choose $\beta=3/2$. Notice that this power law describes the behavior of the distribution we consider in the main text for our example with $\alpha=\gamma=1$ and $\beta=3/2$.
The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
\begin{equation}
  \bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
\end{equation}

where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
\begin{equation}
  \langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
\end{equation}
with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\beta=3/2$ we have 
\begin{equation}
  \frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
\end{equation}
and therefore $\Lambda=N^2$. We have for the mean
\begin{equation}
  \langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
  \label{eq: SM-sample-mean}
\end{equation}
The result is plotted in Fig.~\ref{fig: SM-cut-off}, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

\begin{figure}[h!]
  \includegraphics[width=.475\textwidth]{cut-off-justification.pdf}
  \caption{Analytical prediction for the sample mean in Eq.~\eqref{eq: SM-sample-mean} as a function of the number of extractions, $N$, compared with the results from random extractions (dots).
  }
  \label{fig: SM-cut-off}
\end{figure}

\section{HETEROGENEOUS EXPONENTS}
\label{appendix b}
In this appendix we explore the robustness of our findings in the case in which the exponents characterizing the dynamics of each degree of freedom $x_i$ are not identical for all $i$.

Consider the system
\begin{equation}
  \dot{x}_i = x_i^{\alpha_i} - \sum_jA_{ij}x_i^{\beta_i}x_j^{\gamma_i}
\end{equation}
where the exponents $\alpha_i$, $\beta_i$, $\gamma_i$, are extracted from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and with the same standard deviation $\sigma_e$. The interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
Our results are robust when $\sigma_e$ is small enough and we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
As an example, we show in the plot in Fig.~\ref{fig: SM-het-exp} the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.

\begin{figure}[h!]
  \includegraphics[width=.475\textwidth]{stoch-exp.pdf}
  \caption{Probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
  }
  \label{fig: SM-het-exp}
\end{figure}
